Sufficient Conditions for Reachability in Automata Networks with Priorities

### 2014-03-20 parts/example.tex
several fixes; first final version
titles
link
other chages after review (references, context)
change documentclass to elsarticle + .cls and .bst files
restructuration
simplifications in flattening + begin example

\section{Examples}\label{sec:example}

\towrite{Exemple \texttt{livelock\_and.ph} : différent avec / sans priorités}

\subsection{Large-scale application}

@@

The Boolean model has been automatically encoded into a Process Hitting with 2 classes of priority%
\footnote{Model and scripts are available at
\url{http://www.irccyn.ec-nantes.fr/~folschet/underapprox-tcrsig94.tbz2}.}.
Then, we verified the reachability for the independent activation of 4 outputs of the signalling
cascade (SRE, AP1, NFkB, NFAT) from all possible input combinations (CD45, CD28, TCRlib) using our
new reachability under-approximation (answering either \emph{yes} or \emph{inconclusive}) and a 
previously defined reachability over-approximation \cite{PMR12-MSCS} (answering either \emph{no} or
\emph{inconclusive}).
All result in conclusive decisions, and the under-approximation has been satisfied in 12 cases (over
32) proving the satisfiability of the concerned reachability property in the encoded Boolean network
(and non-satisfiability in the other cases).

>>>>

\section{Biological Examples}\label{sec:example}

\subsection{\ldots (metazoan)}

\todo{Describe the model}

\begin{example}
  The model of \pref{fig:metazoan-php} gives an example of PH with $3$ classes of priorities.
  Although the actions of the highest class of priority are only used to update a cooperative sort,
  the actions of the intermediate priority do not follow the constraints of \pref{ssec:hypothesis},
  thus having a biological meaning by representing different reaction speeds.

  \begin{figure}[htp]
    \centering
    \scalebox{1}{
    \begin{tikzpicture}
      \TSort{(0,4)}{c}{2}{l}
      \TSort{(1,1)}{f}{2}{l}
      \TSort{(6,4)}{a}{2}{r}

      \TSetTick{fc}{0}{00}
      \TSetTick{fc}{1}{01}
      \TSetTick{fc}{2}{10}
      \TSetTick{fc}{3}{11}
      \TSort{(4,1)}{fc}{4}{r}

      \THit{fc_2}{prio}{a_0}{.west}{a_1}
      \path[bounce, bend left]
        \TBounce{a_0}{prio}{a_1}{.south west};

      \THit{c_1}{prio}{a_1}{.west}{a_0}
      \path[bounce, bend right]
        \TBounce{a_1}{prio}{a_0}{.north west};

      \THit{f_1}{bend left=30, in=90}{c_0}{.west}{c_1}
      \path[bounce, bend left]
        \TBounce{c_0}{}{c_1}{.south west};

      \THit{c_1}{selfhit}{c_1}{.west}{c_0}
      \path[bounce, bend right]
        \TBounce{c_1}{}{c_0}{.north west};

      \path (1.8, 1.5) edge[superprio,coopupdate] (3.2, 2);
      \path (0.8, 4.5) edge[superprio,coopupdate] (3.2, 3);
    \end{tikzpicture}
    }
    \caption{
    \label{fig:metazoan-php}
      An example of PH $\oPH$ with $3$ classes of priorities.
      Actions of $\ov{\PHh}^{(3)}$ (lowest priority) are in thin lines,
      actions of $\ov{\PHh}^{(2)}$ (intermediate priority) are in thick lines,
      and actions of $\ov{\PHh}^{(1)}$ (highest priority), whose only purpose is to update the cooperative sort $fc$,
      are not represented but modeled by the double dashed arrows.
    }
  \end{figure}
\end{example}

\begin{example}
  \pref{fig:metazoan-ph} gives the flattened model of the PH of \pref{fig:metazoan-php}.
  It contains $2$ classes of priority, and prioritized actions are only used to update cooperative sorts.
  The simplifications mentioned at the end of \pref{ssec:flattening} have been applied:
  \begin{itemize}
    \item The auto-action $h_1 = \PHhit{c_1}{c_1}{c_0}$ in $\oPH$ has been flattened to a regular action $\PHhit{a_0}{c_1}{c_0}$
      because its playability property could be simplified to $\Fop{h_1} \equiv c_1 \wedge a_0 \equiv a_0$.
    \item The existing cooperative sort $fc$ was kept as in $\oPH$ to avoid an exploding complexity due to
      the numerous updating actions in $\ov{\PHh}^{(1)}$.
  \end{itemize}

  \begin{figure}[htp]
    \centering
    \scalebox{1}{
    \begin{tikzpicture}
      \TSort{(-5,4)}{c}{2}{l}
      \TSort{(0,1)}{f}{2}{l}
      \TSort{(5,4)}{a}{2}{r}

      \TSetTick{fc}{0}{00}
      \TSetTick{fc}{1}{01}
      \TSetTick{fc}{2}{10}
      \TSetTick{fc}{3}{11}
      \TSort{(3,0)}{fc}{4}{r}

      \TSetTick{fa}{0}{00}
      \TSetTick{fa}{1}{01}
      \TSetTick{fa}{2}{10}
      \TSetTick{fa}{3}{11}
      \TSort{(-3,0)}{fa}{4}{l}

      \THit{fc_2}{}{a_0}{.west}{a_1}
      \path[bounce, bend left]
        \TBounce{a_0}{}{a_1}{.south west};

      \THit{fa_3}{}{c_0}{.east}{c_1}
      \path[bounce, bend right]
        \TBounce{c_0}{}{c_1}{.south east};

      \THit{c_1.north east}{}{a_1}{.west}{a_0}
      \path[bounce, bend right]
        \TBounce{a_1}{}{a_0}{.north west};

      \THit{a_0.south west}{}{c_1}{.east}{c_0}
      \path[bounce, bend left]
        \TBounce{c_1}{}{c_0}{.north east};

      \path (0.8, 1.5) edge[prio,coopupdate] (2.2, 1.5);
      \path (-4.3, 4.5) edge[prio,coopupdate] (2.2, 2.5);

      \path (-0.8, 1.5) edge[prio,coopupdate] (-2.2, 1.5);
      \path (4.3, 4.5) edge[prio,coopupdate] (-2.2, 2.5);
    \end{tikzpicture}
    }
    \caption{
    \label{fig:metazoan-ph}
      An example of PH $\PH$ with $2$ classes of priorities.
      Actions of $\PHh^{(2)}$ (unprioritized) are in thin lines,
      and actions of $\PHh^{(1)}$ (prioritized) which are used to update cooperative sorts,
      are abstracted by thick dashed arrows.
    }
  \end{figure}
\end{example}

\subsection{Large-scale Application}

@@

The Boolean model has been automatically encoded into a Process Hitting with 2 classes of priority%
\footnote{Model and scripts are available at
\url{http://www.irccyn.ec-nantes.fr/~folschet/underapprox-tcrsig94.zip}.}.
Then, we verified the reachability for the independent activation of 4 outputs of the signalling
cascade (SRE, AP1, NFkB, NFAT) from all possible input combinations (CD45, CD28, TCRlib) using our
new reachability under-approximation (answering either \emph{yes} or \emph{inconclusive}) and a 
previously defined reachability over-approximation \cite{PMR12-MSCS} (answering either \emph{no} or
\emph{inconclusive}).
All result in conclusive decisions, and the under-approximation has been satisfied in 12 cases (over
32) proving the satisfiability of the concerned reachability property in the encoded Boolean network
(and non-satisfiability in the other cases).

### 2014-03-20 parts/models-ph-php.tex
simplifications in flattening + begin example

In \pref{def:fop}, we finally define the the operator $\Fopsymbol$ which characterizes the playability of an action
given the semantics of PH (\pref{def:play}).
\begin{definition}[Playability property of an action ($\Fopsymbol : \PHh \rightarrow \F$)]\label{def:fop}
  $$\forall h \in \ov{\PHh}, \Fop{h} \equiv \hitter{h} \wedge
    \left( \bigwedge_{\substack{g \in \PHh^{(n)}\\n < \prio(h)}}
    \neg \left( \target{g} \wedge \hitter{g}\right) \right)$$
\end{definition}
%
\begin{theorem}
\label{th:ppplay}
  $h \in \ov{\PHh}$ is playable in $s \in \ov{\PHl}$ if and only if $\Fsem{\Fop{h}}{s}$.
\end{theorem}
%
\newproof{proofppplay}{Proof of \pref{th:ppplay}}
\begin{proofppplay}
  By definition of the semantics of a PH given in \pref{def:play}.
\end{proofppplay}

@@

\todo{Glue up}
\begin{theorem}
\label{th:ppplaysubset}
  $h \in \ov{\PHh}$ is playable in $s \in \ov{\PHl}$ if and only if
  $\exists \mysigma \subseteq s, \Fsem{\Fop{h}}{\mysigma}$.
%  \[\exists i \in \segm{1}{\n}, \exists \mysigma \in \bigtimes{b \in \PHdep{i}{h}} \PHl_b,
%    \mysigma \subseteq s \wedge \Fsem{\F(h)}{\mysigma} \enspace.\]
\end{theorem}
%
\newproof{proofppplaysubset}{Proof of \pref{th:ppplaysubset}}
\begin{proofppplaysubset}
  ($\Rightarrow$) If $\Fsem{\Fop{h}}{s}$ then $\Fop{h} \not\equiv \bot$ and, by property of a DNF,
    at least one of the $\n$ conjunctions of $\Fop{h}$ is true in $s$.
    Suppose the $i$\textsuperscript{th} conjunction is true in $s$, with $i \in \segm{1}{\n}$;
    then we have: $\forall j \in \segm{1}{\m}, p_{i,j} \in s$.
%    Let $\mysigma \in \bigtimes{b \in \PHdep{i}{h}} \PHl_b$
    Let $\mysigma \in \PHsubl_{\PHdep{i}{h}}$
    with $\forall b \in \PHdep{i}{h}, \PHget{\mysigma}{b} = \PHget{s}{b}$.
    We immediately have: $\mysigma \subseteq s$,
    and, by construction of $\PHdep{i}{h}$, $\Fsem{\Fop{h}}{\mysigma}$.

@@

\begin{theorem}[$(\ov{\PHs}; \ov{\PHl}; \ov{\PHa}^{\angles{k}}) \approx \PHflat(\ov{\PHs}; \ov{\PHl}; \ov{\PHa}^{\angles{k}})$]
\label{th:bisimPHP}
  Let $\ov{\PH} = (\ov{\PHs}; \ov{\PHl}; \ov{\PHa}^{\angles{k}})$ a PH with $k$ classes of priorities,
  and $\PH = \PHflat(\ov{\PHs}; \ov{\PHl}; \ov{\PHa}^{\angles{k}}) = (\PHs; \PHl; \PHa^{\angles{2}})$ its flattening.
  \begin{enumerate}
    \item \label{php2ph} $\forall \os, \os' \in \ov{\PHl}$,
      $\os \PHPtrans[\oPH] \os' \Longrightarrow \flats{\os} \PHPtrans[\oPH]^* \flats{\os'}$,
      where $\PHPtrans[\oPH]^*$ is a finite sequence of $\PHPtrans[\oPH]$ transitions.
    \item \label{ph2php} $\forall s, s' \in \PHl$,
      $s \PHPtrans s' \Longrightarrow \unflats{s} = \unflats{s'} \vee
      \unflats{s} \PHPtrans \unflats{s'} \enspace.$
  \end{enumerate}
\end{theorem}
%
\newproof{proofbisimPHP}{Proof of \pref{th:bisimPHP}}
\begin{proofbisimPHP}
  (\ref{php2ph}) Let $\flats{\os} = s$.
    Given the dynamics of a PH (\pref{def:play}), if $\os \PHPtrans[\PH] \os'$,
    then there exists $h \in \ov{\PHh}$ playable in $\os$;
    therefore, $\os' = \os \PHplay h$ and $\target{h} \in \os$.
    Furthermore, given \pref{th:ppplay}, $\Fsem{\Fop{h}}{\os}$,
    and given \pref{th:ppplaysubset}, there exists $\mysigma \subseteq s$ so that $\Fsem{\Fop{h}}{\mysigma}$.
    Therefore, by construction of $\PH$ (\pref{def:flattening}) there exists
    $g = \PHhit{f^{h,i}_\mysigma}{\target{h}}{\bounce{h}} \in \PHh^{(2)}$.
    By construction of $s$ (\pref{def:flattening}), $\PHget{s}{f^{h,i}} = f^{h,i}_\mysigma$ and $\target{h} = \target{g} \in s$.
    Therefore, $g$ is playable in $s$.
    In $s \PHplay g$, the only playable actions are those in $\PHh^{(1)}$ having $\bounce{h} = \bounce{g}$ as hitter
    and updating cooperative sorts, allowing to reach $\flats{\os'}$ in a finite number of actions.
    Thus, $\flats{\os} \PHPtrans[\oPH]^* \flats{\os'}$.

  (\ref{ph2php}) Let $\os = \unflats{s}$.
    Given the dynamics of a PH (\pref{def:play}), if $s \PHPtrans s'$,
    then there exists $g \in \PHh$ playable in $s$; therefore, $s' = s \PHplay g$ and $\target{g} \in s$.
    If $\prio(g) = 1$ then only the active process of a cooperative sort has evolved, and $\unflats{s} = \unflats{s'}$.
    Otherwise, $\prio(g) = 2$; we note in this case: $g = \PHhit{f^{h,i}_\mysigma}{b_j}{b_k}$.
    By construction of the flattening (\pref{def:flattening}), there exists $h \in \ov{\PHh}$ so that
    $\target{h} = b_j$, $\bounce{h} = b_k$ and $\Fsem{\Fop{h}}{\mysigma}$.
    If $g$ is playable, this means that no other action in $\PHh^{(1)}$ is playable, and especially the cooperative sort
    $f^{h,i}$ is already updated; therefore, $\mysigma \subseteq \os$.
    By \pref{th:ppplaysubset}, it comes that $h$ is playable in $\os$.
    Thus, $\unflats{s} \PHPtrans \unflats{s'}$.
\end{proofbisimPHP}

>>>>

In \pref{def:fop}, we finally define the the operator $\Fopsymbol$ which characterizes the playability of an action
given the semantics of PH (\pref{def:play}).
\begin{definition}[Playability property of an action ($\Fopsymbol : \PHh \rightarrow \F$)]\label{def:fop}
  $$\forall h \in \ov{\PHh}, \Fop{h} \equiv \hitter{h} \wedge
    \left( \bigwedge_{\substack{g \in \PHh^{(n)}\\n < \prio(h)}}
    \neg \left( \target{g} \wedge \hitter{g}\right) \right)$$
\end{definition}
%
\begin{theorem}
\label{th:ppplay}
  $h \in \ov{\PHh}$ is playable in $s \in \ov{\PHl}$ if and only if $\Fsem{\Fop{h}}{s} \wedge \target{h} \in s$.
\end{theorem}
%
\newproof{proofppplay}{Proof of \pref{th:ppplay}}
\begin{proofppplay}
  By definition of the semantics of a PH given in \pref{def:play}.
\end{proofppplay}

@@

\todo{Glue up}
\begin{theorem}
\label{th:ppplaysubset}
  $h \in \ov{\PHh}$ is playable in $s \in \ov{\PHl}$ if and only if
  $\exists \mysigma \subseteq s, \Fsem{\Fop{h}}{\mysigma} \wedge \target{h} \in s$.
%  \[\exists i \in \segm{1}{\n}, \exists \mysigma \in \bigtimes{b \in \PHdep{i}{h}} \PHl_b,
%    \mysigma \subseteq s \wedge \Fsem{\F(h)}{\mysigma} \enspace.\]
\end{theorem}
%
\newproof{proofppplaysubset}{Proof of \pref{th:ppplaysubset}}
\begin{proofppplaysubset}
  ($\Rightarrow$) If $\Fsem{\Fop{h}}{s}$ then $\Fop{h} \not\equiv \bot$ and, by property of a DNF,
    at least one of the $\n$ conjunctions of $\Fop{h}$ is true in $s$.
    Suppose the $i$\textsuperscript{th} conjunction is true in $s$, with $i \in \segm{1}{\n}$;
    then we have: $\forall j \in \segm{1}{\m}, p_{i,j} \in s$.
%    Let $\mysigma \in \bigtimes{b \in \PHdep{i}{h}} \PHl_b$
    Let $\mysigma \in \PHsubl_{\PHdep{i}{h}}$
    with $\forall b \in \PHdep{i}{h}, \PHget{\mysigma}{b} = \PHget{s}{b}$.
    We immediately have: $\mysigma \subseteq s$,
    and, by construction of $\PHdep{i}{h}$, $\Fsem{\Fop{h}}{\mysigma}$.

@@

\begin{theorem}[$(\ov{\PHs}; \ov{\PHl}; \ov{\PHa}^{\angles{k}}) \approx \PHflat(\ov{\PHs}; \ov{\PHl}; \ov{\PHa}^{\angles{k}})$]
\label{th:bisimPHP}
  Let $\ov{\PH} = (\ov{\PHs}; \ov{\PHl}; \ov{\PHa}^{\angles{k}})$ a PH with $k$ classes of priorities,
  and $\PH = \PHflat(\ov{\PHs}; \ov{\PHl}; \ov{\PHa}^{\angles{k}}) = (\PHs; \PHl; \PHa^{\angles{2}})$ its flattening.
  \begin{enumerate}
    \item \label{php2ph} $\forall \os, \os' \in \ov{\PHl}$,
      $\os \PHPtrans[\oPH] \os' \Longrightarrow \flats{\os} \PHPtrans[\oPH]^* \flats{\os'}$,
      where $\PHPtrans[\oPH]^*$ is a finite sequence of $\PHPtrans[\oPH]$ transitions.
    \item \label{ph2php} $\forall s, s' \in \PHl$,
      $s \PHPtrans s' \Longrightarrow \unflats{s} = \unflats{s'} \vee
      \unflats{s} \PHPtrans \unflats{s'} \enspace.$
  \end{enumerate}
\end{theorem}
%
\newproof{proofbisimPHP}{Proof of \pref{th:bisimPHP}}
\begin{proofbisimPHP}
  (\ref{php2ph}) Let $\flats{\os} = s$.
    Given the dynamics of a PH (\pref{def:play}), if $\os \PHPtrans[\PH] \os'$,
    then there exists $h \in \ov{\PHh}$ playable in $\os$;
    therefore, $\os' = \os \PHplay h$.
    Furthermore, given \pref{th:ppplay}, $\Fsem{\Fop{h}}{\os}$ and $\target{h} \in \os$,
    and given \pref{th:ppplaysubset}, there exists $\mysigma \subseteq s$ so that $\Fsem{\Fop{h}}{\mysigma}$.
    Therefore, by construction of $\PH$ (\pref{def:flattening}) there exists
    $g = \PHhit{f^{h,i}_\mysigma}{\target{h}}{\bounce{h}} \in \PHh^{(2)}$.
    By construction of $s$ (\pref{def:flattening}), $\PHget{s}{f^{h,i}} = f^{h,i}_\mysigma$ and $\target{h} = \target{g} \in s$.
    Therefore, $g$ is playable in $s$.
    In $s \PHplay g$, the only playable actions are those in $\PHh^{(1)}$ having $\bounce{h} = \bounce{g}$ as hitter
    and updating cooperative sorts, allowing to reach $\flats{\os'}$ in a finite number of actions.
    Thus, $\flats{\os} \PHPtrans[\oPH]^* \flats{\os'}$.

  (\ref{ph2php}) Let $\os = \unflats{s}$.
    Given the dynamics of a PH (\pref{def:play}), if $s \PHPtrans s'$,
    then there exists $g \in \PHh$ playable in $s$; therefore, $s' = s \PHplay g$ and $\target{g} \in s$.
    If $\prio(g) = 1$ then only the active process of a cooperative sort has evolved, and $\unflats{s} = \unflats{s'}$.
    Otherwise, $\prio(g) = 2$; we note in this case: $g = \PHhit{f^{h,i}_\mysigma}{b_j}{b_k}$.
    By construction of the flattening (\pref{def:flattening}), there exists $h \in \ov{\PHh}$ so that
    $\target{h} = b_j$, $\bounce{h} = b_k$ and $\Fsem{\Fop{h}}{\mysigma}$.
    If $g$ is playable, this means that no other action in $\PHh^{(1)}$ is playable, and especially the cooperative sort
    $f^{h,i}$ is already updated; therefore, $\mysigma \subseteq \os$.
    Furthermore, $b_j \in \os$.
    Thus, by \pref{th:ppplaysubset}, it comes that $h$ is playable in $\os$,
    and $\unflats{s} \PHPtrans \unflats{s'}$.
\end{proofbisimPHP}

We showed in this subsection that it is possible to model a PH model with $k$ classes of priorities as a PH model
with $2$ classes of priorities that respects all Criteria of \pref{ssec:hypothesis} and with the same dynamics.
We note however that the translation given in this section may not be minimal in terms of number of actions
or cooperative sorts.
We give here several ways to simplify the resulting model, but we do not give the proofs.

\paragraph{Additional properties}
It is possible to obtain a flattened PH with the same dynamics
by simplifying the playability property of an action $h = \PHhit{a_i}{b_j}{b_k}$ with the following properties:
\begin{itemize}
  \item it is always possible to simplify the target because its presence is checked besides: $b_j \equiv \top$;
  \item any process $b_l \neq b_j$ of the same sort than $b_k$ always prevents the playability, thus: $b_l \equiv \bot$;
  \item if $c_p, c_q$ are different processes ($c_p \neq c_q$) of the same sort $c$, then $c_p \wedge c_q \equiv \bot$.
\end{itemize}

\paragraph{Removal of the resulting cooperative sorts}
There are two cases where it is possible to remove a cooperative sort $f^{h,i}$ in the flattened PH:
\begin{itemize}
  \item if $\Fop{h} \equiv \top$, then the action $h$ can be modeled in the flattened model by an auto-action
    (as it is always playable if the target is present);
  \item if the $i$\textsuperscript{th} conjunction of $\Fop{h}$ consists of only one element $p$,
    then $h$ can be modeled as a regular action $\PHhit{p}{\target{h}}{\bounce{h}}$ in the flattened PH
    (as, apart from the target, only one process is required).
\end{itemize}

\paragraph{Keeping initial cooperations}
The initial model $\oPH$ may contain cooperative sorts similar than those described in \pref{ssec:hypothesis},
only hit by a specific class of priorities, which is the highest.
It is then possible to simplify the translation by keeping these cooperative sorts and ignoring all updating
actions in the playability properties of the other actions.
In this way, the meaning of these cooperative actions is kept, and they end up modeled as regular cooperative sorts
in the flattened model.
